Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-08-19 main.tex
Added figure illustrating transformation geometry, applied some comments from reviewer.

%\authorinfo{Name1}
           %{Affiliation1}
           %{Email1}
%\authorinfo{Name2\and Name3}
           %{Affiliation2/3}
           %{Email2/3}
\authorinfo{Andrew Wagner\and Yi Ma}
			{University of Illinois at Urbana-Champaign}
			{\{awagner,yima\}@illinois.edu}
\authorinfo{Victor Shia\and Allen Yang\and Mark Murphy}
			{University of California Berkeley}
			{\{vshia,yang,mjmurphy\}@eecs.berkeley.edu}
%\authorinfo{Allen Yang}
			%{University of California Berkeley}
			%{yang@eecs.berkeley.edu}
%\authorinfo{Mark Murphy}
			%{University of California Berkeley}
			%{mjmurphy@eecs.berkeley.edu}
%\authorinfo{Yi Ma}
			%{University of Illinois at Urbana-Champaign}
			%{yima@illinois.edu}

@@

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
For very large problems, the GPU is faster due to higher memory bandwidth, while 
for problems that fit in the larger CPU L3 cache, the CPU is faster.
On both architectures, the proposed implementations significantly outperform
naive library-based implementations, as well as previous systems. 
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}
\vspace{-0.1in}

\section{Introduction} 
\vspace{-0.06in}
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the $m$ pixels of the $n_i$ training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{m\times n_1}, \cdots, A_K\in\Re^{m\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{m\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^m$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

Finally, we present extensive benchmarks to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more will likely require computation to be
distributed over a cluster or network of computers. These systems are not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} \vspace{-0.06in}
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}
\vspace{-0.06in}
In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

\vspace{-0.06in}
\section{Hardware Concurrency} \label{sec:concurrency}
\vspace{-0.06in}
In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors and a PCIe card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server (or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.
%\footnote{Non-shared memory configurations include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

\subsection{CPU Hardware Concurrency}
\vspace{-0.06in}
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds. The baseline architecture for our experiments 
is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{A visual comparison of the algorithm working set to the CPU and GPU
caches.  Note: Although the aspect ratio could not be preserved, all arrays and
caches are drawn so that the area is proportional to the size of the data. } \vspace{-0.1in}
\end{figure}

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.06in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SM), each of which is roughly analogous to a CPU core.
For Fermi architecture GPUs there are up to 16 SMs, and each SM is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

\section{Parallelism in the Face Alignment Stage}\vspace{-0.06in}
\label{sec:alignment}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial estimate of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \shrink\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{CPU Implementation} \vspace{-0.06in}
\label{sec:alignment_implementation_cpu}

@@

\subsection{GPU Implementation} \vspace{-0.06in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:gpu_caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

@@

\section{Parallelization of the Face Recognition Stage} \vspace{-0.06in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients in $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} \vspace{-0.06in}
Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} \vspace{-0.06in}
\label{sec:experiment} 
In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.
%To generate a meaningful comparison between different hardware architectures,
%in all cases we compare the performance of hardware architectures on a
%per-board basis:  For CPU implementations, the benchmark makes use of all of
%the cores in as many CPU's are present.  For GPU implementations, the benchmark
%makes the best use of the entire GPU chip (most GPU boards have a single GPU
%chip).

@@

\begin{figure}[t!]
\centering
\includegraphics[width=3.3in]{figures/alignment_runtime_graph}\vspace{0.06in}
\caption{Alignment stage runtime vs. size of training database.} \vspace{-0.1in}
\label{fig:alignment_stage_runtime}
\end{figure}

>>>>

\authorinfo{Name1}
		   {Affiliation1}
		   {Email1}
\authorinfo{Name2\and Name3}
		   {Affiliation2/3}
		   {Email2/3}
%\authorinfo{Andrew Wagner\and Yi Ma}
			%{University of Illinois at Urbana-Champaign}
			%{\{awagner,yima\}@illinois.edu}
%\authorinfo{Victor Shia\and Allen Yang\and Mark Murphy}
			%{University of California Berkeley}
			%{\{vshia,mjmurphy,yang\}@eecs.berkeley.edu}

@@

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
For very large problems, the GPU is faster due to higher memory bandwidth, while 
for problems that fit in the larger CPU L3 cache, the CPU is faster.
On both architectures, the proposed implementations significantly outperform
naive library-based implementations, as well as previous systems. 
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the $m$ pixels of the $n_i$ training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{m\times n_1}, \cdots, A_K\in\Re^{m\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{m\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^m$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$.  
{\bf Several
classifiers based on $\xx$ have been proposed that all have nearly identical 
performance.  In this paper classification is performed using the following residual:
\begin{equation}
i^* = \argmin{i} || \bb - \ee - A_i \xx_i||_2
\label{eq:classification}
\end{equation}
}
% changed

@@

Finally, we present extensive benchmarks to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more will likely require computation to be
distributed over a cluster or network of computers. These systems are not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}
In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

\section{Hardware Concurrency} \label{sec:concurrency}
In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors and a PCIe card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server (or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.
%\footnote{Non-shared memory configurations include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.

\subsection{CPU Hardware Concurrency}
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds. The baseline architecture for our experiments 
is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{A visual comparison of the algorithm working set to the CPU and GPU
caches.  Note: Although the aspect ratio could not be preserved, all arrays and
caches are drawn so that the area is proportional to the size of the data. } 
\end{figure}

@@

\subsection{GPU Hardware Concurrency} 
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SM), each of which is roughly analogous to a CPU core.
For Fermi architecture GPUs there are up to 16 SMs, and each SM is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}

@@

{\bf
Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that maps pixels in a canonical frame, where the face is well aligned,
back to pixels in the original image.  
This is illustrated in Figure \ref{fig:transformation}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/transformation}
\caption{The simililarity transformation $\tau$ maps pixels in a canonical frame
where the image is aligned back to pixel locations in the original-sized images.}
\label{fig:transformation}
\end{figure}
The training images are prepared as
follows: canonical outer-eye corner locations are defined in the canonical
frame, and the outer-eye corners are manually selected in the training images.
The training images are warped/resampled by a similarity transformation
(rotation, translation, and scale) between the two reference frames that is
computed from the eye corners.
The canonical eye corners are also illustrated in Figure \ref{fig:transformation}

During iterative alignment, the transformation $\tau$ between the original test
image and the canonical frame is initialized via a face detector, and then updated
iteratively.  A different $\tau_i$ is computed separately for each training class $A_i$.  
$\ell_1$ minimization is performed on canonical-sized images that
have been stacked into the columns of matrices.  
Directly solving \eqref{eq:l1min_alignment} in its original formulation is difficult since 
it is a non-convex problem and may exhibit local minima.
Minimization is therefore performed iteratively, alternating between updating
a linearized version of the problem at each $j$th step, and solving the linearized problem:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \shrink\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}

@@

\subsection{GPU Implementation} 
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:gpu_caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

@@

\section{Parallelization of the Face Recognition Stage} 
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients in $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} 
Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} 
\label{sec:experiment} 
In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.
%To generate a meaningful comparison between different hardware architectures,
%in all cases we compare the performance of hardware architectures on a
%per-board basis:  For CPU implementations, the benchmark makes use of all of
%the cores in as many CPU's are present.  For GPU implementations, the benchmark
%makes the best use of the entire GPU chip (most GPU boards have a single GPU
%chip).

@@

\begin{figure}[t!]
\centering
\includegraphics[width=3.35in]{figures/alignment_runtime_graph}
\caption{Alignment stage runtime vs. size of training database.} 
\label{fig:alignment_stage_runtime}
\end{figure}
